\textbf{Advection versus diffusion in intraventricular transport}

### 2025-03-26 paper/main.tex
add div(u) -> stability references, edit supplementaries

\subsection{Numerical approximation and computational model parameters}
We modeled motile cilia as a stress
exerted tangentially on the CSF at the ventricular walls
while employing a slip boundary condition,
motivated by experimental observations of the
flow field in zebrafish embryo brain ventricles, both previously and in this work, in which
the fluid velocity increases towards the
wall~(\Crefpanels{fig:fig1}c)~\cite{Olstad2019CiliaryDevelopment}. 
The Brezzi-Douglas-Marini and discontinuous Galerkin scheme
$\mathrm{BDM}_1$-$\mathrm{DG}_0$
we used to discretize the governing equations of the flow
is computationally more expensive
than more conventional approaches such as a
Taylor-Hood discretization
($P_k-P_{k-1}$ elements)~\cite{Stenberg1990ErrorProblem}, or lower-order enriched
elements, such as the mini-element
($P_1-P_1$ with a velocity stabilization)~\cite{Brezzi2011MixedMethods}.
Here, $P_k$ is the space of continuous Lagrange polynomials of degree less
than or equal to the integer $k$.
We opted for the BDM-DG scheme because of its exactly-mass-conserving property;
by the definition of the elements, the numerical approximation of
the CSF velocity $\uu_h$ satisfies the continuity equation (\Cref{eq:stokes_eqs_cont}).
This yields $\normlinf{\nabla\cdot\uu_h}=0$ to within solver tolerances,
an important property for the stability of the advection-diffusion
equation\cite{cesmelioglu2022compatible, johnson2009numerical}.
In addition to its favorable mass-conserving property,
the BDM-DG discretization scheme 
we employed for the Stokes equations
lets us impose the impermeability condition $\uu\cdot\nn=0$
strongly without any tailored approach,
since the BDM elements control the normal component of the 
velocity~(\Cref{subsec:appendixA1})~\cite{Brezzi1985TwoProblems}.
If instead (for example) Taylor-Hood
elements were to be used, the impermeability boundary condition must
be handled with for example a Nitsche method~\cite{Nitsche1971UberSind}
or with a Lagrange multiplier~\cite{Babuska1973TheMultipliers,
Bertoluzza2017BoundaryHemodynamics}, which affects
convergence properties and increases problem complexity.

@@

\textbf{Only-cilia/no-cardiac model.} This model version has cilia forces
as the only mechanism driving flow. We consider the
Stokes equations~\Cref{eq:appendix_stokes_eqs} with the slip boundary conditions
\begin{subequations}
    \begin{alignat}{2}
        \uu\cdot\nn &= 0 \quad &&\mathrm{on} \ \Gamma=\Gc\cup\Gs,
        \label{eq:modelA_slip_BC_impermeability} \\
        \bsigpar &= \btau \quad &&\mathrm{on} \ \Gc,
        \label{eq:modelA_slip_BC_shear} \\
        \bsigpar &= \mathbf{0} \quad &&\mathrm{on} \ \Gs,\label{eq:modelA_full_slip_BC}
    \end{alignat}\label{eq:modelA_slip_BC}%
\end{subequations}%
on the boundary $\Gamma=\Gc\cup\Gs$. The boundary $\Gs$ is a free-slip boundary,
where we impose the free-slip condition~\Cref{eq:modelA_full_slip_BC}.
The first condition~\Cref{eq:modelA_slip_BC_impermeability} is the
impermeability condition, enforcing no flow normal to the boundary,
whereas~\Cref{eq:modelA_slip_BC_shear} imposes a stress tangential to
the boundary. These tangential stresses are applied to represent
the forces that the collective cilia motion exerts on the cerebrospinal fluid.

The boundary conditions~\Cref{eq:modelA_slip_BC} are set weakly through the stress
boundary integral 
\begin{equation}
    \intG{\bsig\nn\cdot\vv} =  \intG{\hat{\bsig}\cdot\vv}
    \label{eq:boundary_integral_BC1}
\end{equation}
in the weak form~\Cref{eq:modelA_weak_form1} of the Stokes equations.
We decompose the traction vector $\hat{\bsig}$ into a normal and tangential component:
\begin{equation*}
    \hat{\bsig} = \bsigperp + \bsigpar
\end{equation*}
Inserting the above into~\Cref{eq:boundary_integral_BC1},
combined with application of the tangential traction
condition~\Cref{eq:modelA_slip_BC_shear} and the
free-slip condition~\Cref{eq:modelA_full_slip_BC}, yields
\begin{equation}
    \intG{\hat{\bsig}\cdot\vv} = \int_{\Gc\cup\Gs}{(\bsigperp + \bsigpar)\cdot\vv}\, ds = 
    \int_{\Gc\cup\Gs}{\bsigperp\cdot\vv}\, ds
    + \intGc{\btau\cdot\vv} = \intG{\bsigperp\cdot\vv} + \intGc{\btau\cdot\vv}.
    \label{eq:boundary_integral_BC2}
\end{equation}
The second term on the right-hand side of~\Cref{eq:boundary_integral_BC2}
is moved to the right-hand side of the weak form, since $\btau$ is known
 Using $\hat{\bsig}_{\perp} = (\hat{\bsig}\cdot\nn)\nn$,
 we can write the first term on the right-hand side of~\Cref{eq:boundary_integral_BC2} as
\begin{equation}
    \intG{\bsigperp\cdot\vv} = \intG{((\hat{\bsig}\cdot\nn)\nn)\cdot\vv} =
    \intG{(\hat{\bsig}\cdot\nn)(\vv\cdot\nn)}.
    \label{eq:boundary_integral_BC_normal}
\end{equation}
Since $\Gp=\emptyset$ in this model version, and the term~\Cref{eq:boundary_integral_BC_normal}
vanishes owing to the construction $\VV$. As a result, for the cilia-only/no-cardiac model,
the weak form of the momentum equation takes the following form:
\begin{equation*}
    \intO{\rho\pdifft{\uu}\cdot\vv}
    + \intO{2\mu\beps(\uu) : \beps(\vv)}
    - \intO{p\nabla\cdot\vv} = \intGc{\btau\cdot\vv}, \quad\forall\vv\in\VV.
\end{equation*}
Next, multiplying the continuity equation~\Cref{eq:BDM_stokes_div}
with a test function $q\in Q$, we require that
\begin{equation}
    -\intO{q\nabla\cdot\uu} = 0,\quad\forall q\in Q
    \label{eq:pressure_weak_form}
\end{equation}
The full weak form of the Stokes equations for this flow model is then:
find $(\uu, \, p) \in \VV\times Q$, such that
\begin{subequations}
    \begin{alignat}{5}
        &\intO{\rho\pdifft{\uu}\cdot\vv}
        + &\intO{2\mu\beps(\uu) : \beps(\vv)}
        - &\intO{p\nabla\cdot\vv}
        &&= \intGc{\btau\cdot\vv}, &&\quad\forall\vv\in\VV, \label{eq:stokes_weak_mom_modelA}\\ 
        &{} &{}  -&\intO{q\nabla\cdot\uu} &&= 0, &&\quad\forall q\in Q.
        \label{eq:stokes_weak_cont_modelA}
    \end{alignat}%
    \label{eq:stokes_weak_modelA}%
\end{subequations}%
We observe that, in this case with slip boundary conditions on the entirety of
the boundary $\Gamma$, for any real constant $C$ the vector
$(\uu, p)=(\mathbf{0}, C)$ satisfies the equations.
Hence the system~\Cref{eq:stokes_weak_modelA} is singular with a one-dimensional nullspace
of constant pressures.
Consequently, assembly of our discretization of the  equations~\Cref{eq:stokes_weak_modelA}
into a linear equation system $\mathbf{A}\mathbf{x}=\mathbf{b}$ yields a singular matrix $\mathbf{A}$.
To ensure existence of solutions, we orthogonalize the right-hand side vector $\mathbf{b}$ with
respect to the pressure nullspace basis (the set of all constants), before solving the linear
system. To obtain a unique solution, which together with existence renders the
problem well-posed, information of the pressure nullspace is passed to the linear solver.
Let us remark that the singularity is a consequence of the fact that the
impermeability condition is assumed to hold on the \emph{entire} domain boundary
for this model version, which leaves the pressure under-determined as a result of no
boundary condition specifying the pressure.

The tangential traction $\btau = \tau\lambda(\xx) P_{\nn}(\rr)$ that represents the cilia
motion differ for the cilia regions. We heuristically determined the constant $\tau= 0.65$ mPa
and the spatially varying function $\lambda(\xx)$. The definitions of $\rr$ and $\lambda(\xx)$
for the different cilia populations are summarized in~\Cref{tab:r_vector_and_lambda_function}.
\begingroup
\setlength{\tabcolsep}{20pt}
\renewcommand{\arraystretch}{1.5}
\begin{table}[H]
    \centering
    \caption{Definitions of $\rr$ and $\lambda(\xx)$ used in for the cilia tangential 
    traction boundary condition $\btau = \tau\lambda(\xx)P_{\nn}(\rr)$,
    where $\tau=0.65$ mPa. The $x-$coordinates $x_{0, i}$ and $x_{e, i}$ define 
    the cilia domain in the dorsal ($i=d$) and ventral ($i=v$) regions of 
    the middle ventricle.}\label{tab:r_vector_and_lambda_function}
    \begin{tabular}{lcc}
        \toprule
        Cilia population & $\rr$ & $\lambda(\xx)$ \\
        \midrule 
        Anterior region, anterior ventricle 
        & $-(1, 0, 1)$ & $\frac{2}{5}$ \\
        Posterior region, anterior ventricle
        & $\ \ \ (1, 0, 1)$ & 1 \\
        Dorsal region, middle ventricle     
        & $-(1, 0, 1)$
        & $\frac{11}{4}\left(\frac{x-x_{0, d}}{x_{e, d} - x_{0, d}}\right)$\\
        Ventral region, middle ventricle    
        & $\ \ \ (1, 0, 1)$
        & $\frac{1}{2}\left(1 - \frac{x-x_{0, v}}{x_{e, v} - x_{0, v}}\right)$ \\
        \bottomrule
    \end{tabular}
\end{table}
\endgroup

\textbf{Only-cardiac/no-cilia model.} For this model version,
the only physical mechanism inducing fluid flow is the cardiac cycle.
We model the influence of the cardiac cycle with a normal pressure boundary condition:
\begin{subequations}
    \begin{alignat}{2}
      \uu\cdot\nn = 0, \ \hat{\bsig}_{\parallel} &= \mathbf{0} \quad &&\mathrm{on} \ \Gs,
      \label{eq:modelB_slip_BC} \\
      (\mu\nabla\uu - p\mathbf{I})\nn &= \Tilde{p}(t)\nn \quad &&\mathrm{on} \ \Gp,
      \label{eq:modelB_pressure_BC}
    \end{alignat}%
\end{subequations}%
where the boundary $\Gamma = \Gs\cup\Gp$ consists of a free-slip boundary $\Gs$,
and a part $\Gp$ where a normal traction type of boundary condition is imposed.
Note, however, that the transpose velocity gradient term must be
included inside the parentheses on the left-hand side of~\Cref{eq:modelB_pressure_BC}
for it to be the traction $\bsig$. Since this is not the case,
we rather refer to this boundary condition as a normal pressure boundary condition.
The boundary $\Gs$ is handled as discussed in the previous section,
such that the stress boundary integral~\Cref{eq:boundary_integral_BC1}
vanishes over $\Gs$. Therefore, we now have
\begin{align}
    \intG{\hat{\bsig}\cdot\vv} = \intGp{\hat{\bsig}\cdot\vv}
    &= \intGp{(\mu(\nabla\uu + (\nabla\uu)^T) - p\mathbf{I})\nn\cdot\vv}\notag \\
    &= \intGp{\mu(\nabla\uu)^T\nn\cdot\vv} + \intGp{(\mu\nabla\uu
    - p\mathbf{I})\nn\cdot\vv}\notag\\
    &= \intGp{\mu(\nabla\uu)^T\nn\cdot\vv} + \intGp{\Tilde{p}(t)(\vv\cdot\nn)}
    \label{eq:modelB_boundary_integral1}
\end{align}
We leave the first term on the left-hand side
of~\Cref{eq:modelB_boundary_integral1} in the weak form of
the momentum equation. The second term is determined by
the boundary condition~\Cref{eq:modelB_pressure_BC},
and is therefore moved to the right-hand side of the weak form.
The weak form of the momentum equation is then
\begin{equation}
    \begin{aligned}
    \intO{\rho\pdifft{\uu}\cdot\vv} + \intO{2\mu\beps(\uu) : \beps(\vv)} -
    \intO{p\nabla\cdot\vv} &-\intGp{\mu(\nabla\uu)^T\nn\cdot\vv} \\
    &=\intGp{\Tilde{p}(t)(\vv\cdot\nn)}, \quad\forall\vv\in\VV.
    \end{aligned}
    \label{eq:weak_form_mom_eq_modelB}
\end{equation}
Also considering the weak form of the continuity equation,
the weak formulation for the cardiac-only/no-cilia model version
then reads: find $(\uu,\, p)\in\VV\times Q$
such that~\Cref{eq:weak_form_mom_eq_modelB} and~\Cref{eq:pressure_weak_form}
hold for all $(\vv, \, q)\in\VV\times Q$.
To simulate the back-and-forth motion induced by the cardiac cycle observed in
previous work~\cite{Olstad2019CiliaryDevelopment}, we apply sinusoidal forcing.
The boundary $\Gp$ has two separate regions: one located on the
anterior ventricle and one on the posterior ventricle~(\Crefpanels{fig:fig1}{d}, blue markers).
For the normal pressure boundary condition, we set $\Tilde{p}(t) = 0$ on the anterior part and
$\Tilde{p}(t) = p_c(t) = -A\sin{\omega t}$ on the posterior part. The value of the amplitude
$A$ was based on heuristics and $\omega = 2\pi f$, where $f=2.22$ Hz
is the cardiac cycle frequency.

@@

\subsection{Discretization of the Stokes equations}
We will outline the discretization of the weak
formulation~\Cref{eq:abstract_weak_form_modelC,
eq:stokes_bilinear_form,eq:stokes_skew_form,eq:stokes_linear_form}
of the Stokes equations for the baseline model. A similar approach can
be used to discretize the weak formulation for the other model versions.
First, we consider the spatial discretization of the weak formulation.
We rewrite the weak form in terms of the discretized variables,
denoted with the subscript $h$, on the discrete domain $\Omega_h$:
\begin{subequations}
    \begin{alignat}{2}
        \mathcal{A}(\uu_h, \vv_h) + \mathcal{B}(\vv_h, p_h)
        &= \mathcal{L}(\vv_h), &&\quad\forall\vv_h\in\VV_h, \\
        \mathcal{B}(\uu_h, q_h) &= 0, &&\quad\forall q_h\in Q_h.
    \end{alignat}%
    \label{eq:no_stabilization_abstract_discrete_weak_form_modelC}%
\end{subequations}%
Our choice of discretization are finite elements
of the Brezzi-Douglas-Marini (BDM) family~\cite{Brezzi1985TwoProblems, Nedelec1986AIR3}
for the velocity
space and discontinuous Lagrange polynomial elements for the pressure. We refer to the latter
as discontinuous Galerkin (DG) elements. For all cells $K$ of the computational mesh $\Omega_h$,
we define
\begin{align*}
    {\VV}_h &= \Big\{\vv_h : \vv_h\in\mathbf{H}(\mathrm{div});
    \vv_h\vert_K\in\big[P_k\big]^d;
    \forall K\in\Omega_h;
    \vv_h\cdot\nn=0 \ \text{on } \Gamma\setminus\Gp\Big\},\\
    Q_h &= \Big\{q_h : q_h\in L^2;  q_h\vert_K\in P_{k-1}; \forall K\in\Omega_h\Big\},
\end{align*}
where $P_k$ is the space of polynomials of degree less than or equal to $k$.
The BDM family of elements have degrees of freedom
associated with integral moments of vector components normal to facets of the mesh,
and is therefore non-conforming. However, the BDM-DG is $\mathbf{H}(\mathrm{div})$
conforming and thus exactly mass-conserving, because the continuity equation can be satisfied
in every computational mesh cell due to the fact that $\nabla\cdot\VV_h\subset Q_h$.
Moreover, control of normal components of $\uu_h$ conveniently 
lets us enforce the impermeability condition $\uu_h\cdot\nn=0$ strongly.
requires a stabilization of the bilinear form $\mathcal{A}(\uu_h, \vv_h)$.

>>>>

\subsection{Numerical approximation and computational model parameters}
We modeled motile cilia as a stress
exerted tangentially on the CSF at the ventricular walls
while employing a slip boundary condition,
motivated by experimental observations of the
flow field in zebrafish embryo brain ventricles, both previously and in this work, in which
the fluid velocity increases towards the
wall~(\Crefpanels{fig:fig1}c)~\cite{Olstad2019CiliaryDevelopment}. 
The Brezzi-Douglas-Marini and discontinuous Galerkin scheme
$\mathrm{BDM}_1$-$\mathrm{DG}_0$
we used to discretize the governing equations of the flow
is computationally more expensive
than more conventional approaches such as a
Taylor-Hood discretization
($P_k-P_{k-1}$ elements)~\cite{Stenberg1990ErrorProblem}, or lower-order enriched
elements, such as the mini-element
($P_1-P_1$ with a velocity stabilization)~\cite{Brezzi2011MixedMethods}.
Here, $P_k$ is the space of continuous Lagrange polynomials of degree less
than or equal to the integer $k$.
We opted for the BDM-DG scheme because of its exactly-mass-conserving property;
by the definition of the elements, the numerical approximation of
the CSF velocity $\uu_h$ satisfies the continuity equation (\Cref{eq:stokes_eqs_cont}).
This yields $\normlinf{\nabla\cdot\uu_h}=0$ to within solver tolerances,
an important property for the stability of the advection-diffusion
equation\cite{cesmelioglu2022compatible, johnson2009numerical, gresho1998incompressible}.
In addition to its favorable mass-conserving property,
the BDM-DG discretization scheme 
we employed for the Stokes equations
lets us impose the impermeability condition $\uu\cdot\nn=0$
strongly without any tailored approach,
since the BDM elements control the normal component of the 
velocity~(\Cref{subsec:appendixA1})~\cite{Brezzi1985TwoProblems}.
If instead (for example) Taylor-Hood
elements were to be used, the impermeability boundary condition must
be handled with for example a Nitsche method~\cite{Nitsche1971UberSind}
or with a Lagrange multiplier~\cite{Babuska1973TheMultipliers,
Bertoluzza2017BoundaryHemodynamics}, which affects
convergence properties and increases problem complexity.

@@

\textbf{Cilia-driven/no-cardiac model.} This model version has cilia forces
as the only mechanism driving flow. We consider the
Stokes equations~\Cref{eq:appendix_stokes_eqs} with the slip boundary conditions
\begin{subequations}
    \begin{alignat}{2}
        \uu\cdot\nn &= 0 \quad &&\mathrm{on} \ \Gamma=\Gc\cup\Gs,
        \label{eq:modelA_slip_BC_impermeability} \\
        \bsigpar &= \btau \quad &&\mathrm{on} \ \Gc,
        \label{eq:modelA_slip_BC_shear} \\
        \bsigpar &= \mathbf{0} \quad &&\mathrm{on} \ \Gs,\label{eq:modelA_full_slip_BC}
    \end{alignat}\label{eq:modelA_slip_BC}%
\end{subequations}%
on the boundary $\Gamma=\Gc\cup\Gs$. The boundary $\Gs$ is a free-slip boundary,
where we impose the free-slip condition~\Cref{eq:modelA_full_slip_BC}.
The first condition~\Cref{eq:modelA_slip_BC_impermeability} is the
impermeability condition, enforcing no flow normal to the boundary,
whereas~\Cref{eq:modelA_slip_BC_shear} imposes a traction tangential to
the boundary. This tangential traction is applied to represent
the forces that the collective cilia motion exerts on the cerebrospinal fluid.

The boundary conditions~\Cref{eq:modelA_slip_BC} are set weakly through the
boundary integral 
\begin{equation}
    \intG{\bsig\nn\cdot\vv} =  \intG{\hat{\bsig}\cdot\vv}
    \label{eq:boundary_integral_BC1}
\end{equation}
in the weak form~\Cref{eq:modelA_weak_form1} of the Stokes equations.
We decompose the traction vector $\hat{\bsig}$ into a normal and tangential component:
\begin{equation*}
    \hat{\bsig} = \bsigperp + \bsigpar.
\end{equation*}
Inserting the above into~\Cref{eq:boundary_integral_BC1},
combined with application of the tangential traction
condition~\Cref{eq:modelA_slip_BC_shear} and the
free-slip condition~\Cref{eq:modelA_full_slip_BC}, yields
\begin{equation}
    \intG{\hat{\bsig}\cdot\vv} = \int_{\Gc\cup\Gs}{(\bsigperp + \bsigpar)\cdot\vv}\, ds = 
    \int_{\Gc\cup\Gs}{\bsigperp\cdot\vv}\, ds
    + \intGc{\btau\cdot\vv} = \intG{\bsigperp\cdot\vv} + \intGc{\btau\cdot\vv}.
    \label{eq:boundary_integral_BC2}
\end{equation}
Since $\btau$ is known, the second term on the right-hand side of~\Cref{eq:boundary_integral_BC2}
is moved to the right-hand side of the weak form.
With $\hat{\bsig}_{\perp} = (\hat{\bsig}\cdot\nn)\nn$,
we can write the first term on the right-hand side of~\Cref{eq:boundary_integral_BC2} as
\begin{equation}
    \intG{\bsigperp\cdot\vv} = \intG{((\hat{\bsig}\cdot\nn)\nn)\cdot\vv} =
    \intG{(\hat{\bsig}\cdot\nn)(\vv\cdot\nn)}.
    \label{eq:boundary_integral_BC_normal}
\end{equation}
This term vanishes owing to the construction $\VV$.
As a result, for the cilia-driven/no-cardiac model,
the weak form of the momentum equation takes the following form:
\begin{equation*}
    \intO{\rho\pdifft{\uu}\cdot\vv}
    + \intO{2\mu\beps(\uu) : \beps(\vv)}
    - \intO{p\nabla\cdot\vv} = \intGc{\btau\cdot\vv}, \quad\forall\vv\in\VV.
\end{equation*}
Next, multiplying the continuity equation~\Cref{eq:BDM_stokes_div}
with a test function $q\in Q$, we require that
\begin{equation}
    -\intO{q\nabla\cdot\uu} = 0,\quad\forall q\in Q.
    \label{eq:pressure_weak_form}
\end{equation}
The full weak form of the Stokes equations for this flow model is then:
find $(\uu, \, p) \in \VV\times Q$, such that
\begin{subequations}
    \begin{alignat}{5}
        &\intO{\rho\pdifft{\uu}\cdot\vv}
        + &\intO{2\mu\beps(\uu) : \beps(\vv)}
        - &\intO{p\nabla\cdot\vv}
        &&= \intGc{\btau\cdot\vv}, &&\quad\forall\vv\in\VV, \label{eq:stokes_weak_mom_modelA}\\ 
        &{} &{}  -&\intO{q\nabla\cdot\uu} &&= 0, &&\quad\forall q\in Q.
        \label{eq:stokes_weak_cont_modelA}
    \end{alignat}%
    \label{eq:stokes_weak_modelA}%
\end{subequations}%
We observe that, in this case with slip boundary conditions on the entirety of
the boundary $\Gamma$, for any real constant $C$ the vector
$(\uu, p)=(\mathbf{0}, C)$ satisfies the equations.
Hence the system~\Cref{eq:stokes_weak_modelA} is singular with a one-dimensional nullspace
of constant pressures.
Consequently, assembly of our discretization of the  equations~\Cref{eq:stokes_weak_modelA}
into a linear equation system $\mathbf{A}\mathbf{x}=\mathbf{b}$ yields a singular matrix $\mathbf{A}$.
To ensure existence of solutions, we orthogonalize the right-hand side vector $\mathbf{b}$ with
respect to the pressure nullspace basis (the set of all constants), before solving the linear
system. To obtain a unique solution, which together with existence renders the
problem well-posed, information of the pressure nullspace is passed to the linear solver.
Let us remark that the singularity is a consequence of the fact that the
impermeability condition is assumed to hold on the \emph{entire} domain boundary
for this model version, which leaves the pressure under-determined as a result of no
boundary condition specifying the pressure.

The tangential traction $\btau = \tau\lambda(\xx) P_{\nn}(\rr)$ that represents the cilia
motion differ for the cilia regions. We heuristically determined the constant $\tau= 0.65$ mPa
and the spatially varying function $\lambda(\xx)$. The definitions of $\rr$ and $\lambda(\xx)$
for the different cilia populations are summarized in~\Cref{tab:r_vector_and_lambda_function}.
\begingroup
\setlength{\tabcolsep}{20pt}
\renewcommand{\arraystretch}{1.5}
\begin{table}[H]
    \centering
    \caption{Definitions of $\rr$ and $\lambda(\xx)$ used in the cilia tangential 
    traction boundary condition $\btau = \tau\lambda(\xx)P_{\nn}(\rr)$,
    where $\tau=0.65$ mPa. The $x-$coordinates $x_{0, i}$ and $x_{e, i}$ define 
    the cilia domain in the dorsal ($i=d$) and ventral ($i=v$) regions of 
    the middle ventricle.}\label{tab:r_vector_and_lambda_function}
    \begin{tabular}{lcc}
        \toprule
        Cilia population & $\rr$ & $\lambda(\xx)$ \\
        \midrule 
        Anterior region, anterior ventricle 
        & $-(1, 0, 1)$ & $\frac{2}{5}$ \\
        Posterior region, anterior ventricle
        & $\ \ \ (1, 0, 1)$ & 1 \\
        Dorsal region, middle ventricle     
        & $-(1, 0, 1)$
        & $\frac{11}{4}\left(\frac{x-x_{0, d}}{x_{e, d} - x_{0, d}}\right)$\\
        Ventral region, middle ventricle    
        & $\ \ \ (1, 0, 1)$
        & $\frac{1}{2}\left(1 - \frac{x-x_{0, v}}{x_{e, v} - x_{0, v}}\right)$ \\
        \bottomrule
    \end{tabular}
\end{table}
\endgroup

\textbf{Cardiac-induced/no-cilia model.} For this model version,
the only physical mechanism inducing fluid flow is the cardiac cycle.
The cardiac cycle generates directional back-and-forth motion of the CSF
in the zebrafish brain ventricles.
We model the influence of the cardiac cycle with a normal pressure boundary condition:
\begin{subequations}
    \begin{alignat}{2}
      \uu\cdot\nn = 0, \ \hat{\bsig}_{\parallel} &= \mathbf{0} \quad &&\mathrm{on} \ \Gs,
      \label{eq:modelB_slip_BC} \\
      (\mu\nabla\uu - p\mathbf{I})\nn &= \Tilde{p}(t)\nn \quad &&\mathrm{on} \ \Gp,
      \label{eq:modelB_pressure_BC}
    \end{alignat}%
\end{subequations}%
where the boundary $\Gamma = \Gs\cup\Gp$ consists of a free-slip boundary $\Gs$,
and a part $\Gp$ where a normal traction type of boundary condition is imposed.
Note, however, that the transpose velocity gradient term must be
included inside the parentheses on the left-hand side of~\Cref{eq:modelB_pressure_BC}
for it to be the traction $\bsig$. Since this is not the case,
we rather refer to this boundary condition as a normal pressure boundary condition.
The boundary $\Gs$ is handled as discussed in the previous section,
such that the stress boundary integral~\Cref{eq:boundary_integral_BC1}
vanishes over $\Gs$. Therefore, we now have
\begin{align}
    \intG{\hat{\bsig}\cdot\vv} = \intGp{\hat{\bsig}\cdot\vv}
    &= \intGp{(\mu(\nabla\uu + (\nabla\uu)^T) - p\mathbf{I})\nn\cdot\vv}\notag \\
    &= \intGp{\mu(\nabla\uu)^T\nn\cdot\vv} + \intGp{(\mu\nabla\uu
    - p\mathbf{I})\nn\cdot\vv}\notag\\
    &= \intGp{\mu(\nabla\uu)^T\nn\cdot\vv} + \intGp{\Tilde{p}(t)(\vv\cdot\nn)}.
    \label{eq:modelB_boundary_integral1}
\end{align}
We leave the first term on the left-hand side
of~\Cref{eq:modelB_boundary_integral1} in the weak form of
the momentum equation. The second term is determined by
the boundary condition~\Cref{eq:modelB_pressure_BC},
and is therefore moved to the right-hand side of the weak form.
The weak form of the momentum equation is then
\begin{equation}
    \begin{aligned}
    \intO{\rho\pdifft{\uu}\cdot\vv} + \intO{2\mu\beps(\uu) : \beps(\vv)} -
    \intO{p\nabla\cdot\vv} &-\intGp{\mu(\nabla\uu)^T\nn\cdot\vv} \\
    &=\intGp{\Tilde{p}(t)(\vv\cdot\nn)}, \quad\forall\vv\in\VV.
    \end{aligned}
    \label{eq:weak_form_mom_eq_modelB}
\end{equation}
Also considering the weak form of the continuity equation,
the weak formulation for the cardiac-only/no-cilia model version
then reads: find $(\uu,\, p)\in\VV\times Q$
such that~\Cref{eq:weak_form_mom_eq_modelB} and~\Cref{eq:pressure_weak_form}
hold for all $(\vv, \, q)\in\VV\times Q$.
To simulate the back-and-forth motion induced by the cardiac cycle observed in
previous work~\cite{Olstad2019CiliaryDevelopment}, we apply sinusoidal forcing.
The boundary $\Gp$ has two separate regions: one located on the
anterior ventricle and one on the posterior ventricle~(\Crefpanels{fig:fig1}{d}, blue markers).
For the normal pressure boundary condition, we set $\Tilde{p}(t) = 0$ on the anterior part and
$\Tilde{p}(t) = p_c(t) = -A\sin{\omega t}$ on the posterior part. The value of the amplitude
$A$ was based on heuristics and $\omega = 2\pi f$, where $f=2.22$ Hz
is the cardiac cycle frequency.

@@

\subsection{Discretization of the Stokes equations}
We will outline the discretization of the weak
formulation~\Cref{eq:abstract_weak_form_modelC,
eq:stokes_bilinear_form,eq:stokes_skew_form,eq:stokes_linear_form}
of the Stokes equations for the baseline model. A similar approach can
be used to discretize the weak formulation for the other model versions.
First, we consider the spatial discretization of the weak formulation.
We rewrite the weak form in terms of the discretized variables,
denoted with the subscript $h$, on the discrete domain $\Omega_h$:
\begin{subequations}
    \begin{alignat}{2}
        \mathcal{A}(\uu_h, \vv_h) + \mathcal{B}(\vv_h, p_h)
        &= \mathcal{L}(\vv_h), &&\quad\forall\vv_h\in\VV_h, \\
        \mathcal{B}(\uu_h, q_h) &= 0, &&\quad\forall q_h\in Q_h.
    \end{alignat}%
    \label{eq:no_stabilization_abstract_discrete_weak_form_modelC}%
\end{subequations}%
Our choice of discretization are finite elements
of the Brezzi-Douglas-Marini (BDM) family~\cite{Brezzi1985TwoProblems, Nedelec1986AIR3}
for the velocity
space and discontinuous Lagrange polynomial elements for the pressure. We refer to the latter
as discontinuous Galerkin (DG) elements. For all cells $K$ of the computational mesh $\Omega_h$,
we define
\begin{align*}
    \VV_h &= \Big\{\vv_h : \vv_h\in\mathbf{H}(\mathrm{div});
    \vv_h\vert_K\in\big[P_k(K)\big]^d;
    \forall K\in\Omega_h;
    \vv_h\cdot\nn=0 \ \text{on } \Gamma\setminus\Gp\Big\},\\
    Q_h &= \Big\{q_h : q_h\in L^2;  q_h\vert_K\in P_{k-1}(K); \forall K\in\Omega_h\Big\},
\end{align*}
where $P_k$ is the space of polynomials of degree less than or equal to $k$.
The BDM family of elements have degrees of freedom
associated with integral moments of vector components normal to facets of the mesh.
Because $\VV_h\not\subset\VV$, the elements are non-conforming.
This BDM-DG scheme for the Stokes equations
is $\mathbf{H}(\mathrm{div})$ conforming and thus exactly mass-conserving,
because the continuity equation can be satisfied
in every computational mesh cell due to the fact that $\nabla\cdot\VV_h\subset Q_h$.
Moreover, control of normal components of $\uu_h$ conveniently 
lets us enforce the impermeability condition $\uu_h\cdot\nn=0$ strongly.
requires a stabilization of the bilinear form $\mathcal{A}(\uu_h, \vv_h)$.

Because the elements of $\VV_h$ define the normal components of $\uu_h$,
continuity in tangential components of $\uu_h$ on interior facets of
the mesh is ensured through stabilization
of the discrete weak form~\Cref{eq:no_stabilization_abstract_discrete_weak_form_modelC}.
We apply the stabilization outlined in Hong~\emph{et al.}~\cite{Hong2016AEquations}.
First, we define the average and jump operators, respectively, as
\begin{equation}
    \avg{\uu} = \frac{\uu^+\cdot\nn^+ - \uu^-\cdot\nn^-}{2}
    \qquad\mathrm{and}\qquad \jump{\uu} = \uu^+ - \uu^-,
    \label{eq:operators_BDM}
\end{equation}
where $\uu$ is a first-order tensor and 
$\nn$ is the normal vector on a facet separating two cells,
and the plus and minus signs denote variables in each of the two cells.
The normal vector is chosen such that it points towards the exterior of
its corresponding cell. On the boundary, the operators are defined as
$\avg{\uu} = \uu\cdot\nn$ and $\jump{\uu} = \uu$,
with $\nn$ pointing to the exterior of the domain. Note that average operator
defined in~\Cref{eq:operators_BDM} reduces the rank of the tensor $\uu$ by one,
whereas the jump operator preserves rank. This extends to tensors of higher order.
We now define and add the bilinear form
\begin{align*}
    \mathcal{A}_{\mathrm{stab}}(\uu_h, \vv_h) &=
    \sum_{f\in \mathcal{F}}\int_f{2\mu\frac{\gamma}{\avg{h}}\jump{P_{\nn}(\uu_h)}
    \cdot\jump{P_{\nn}(\vv_h)}}\,\mathrm ds\\
    &-\sum_{f\in \mathcal{F}}\int_f{2\mu\avg{\beps(\uu_h)}\cdot\jump{\vv_h}}\,\mathrm ds
    -\sum_{f\in \mathcal{F}}\int_f{2\mu\avg{\beps(\vv_h)}\cdot\jump{\uu_h}}\,\mathrm ds
\end{align*}
to the discrete weak formulation~\Cref{eq:no_stabilization_abstract_discrete_weak_form_modelC},
where $\mathcal{F}$ is the set of all interior facets of the computational mesh.
The first term of $\mathcal{A}_{\mathrm{stab}}$ penalizes jumps
in the tangential components of the velocity
across interior facets, where the penalty parameter $\gamma>0$ must be chosen large enough
to achieve stability. The penalization term
scales with the average of the cell diameter of neighboring mesh cells,
with $h$ defined as the maximal distance between two points in a cell. The other two terms
of $\mathcal{A}_{\mathrm{stab}}$ are added to ensure consistency and symmetry of
the bilinear form, such that the problem is well-defined.
